---
title: "HW3"
author: "Zhu Xiaohan"
date: "2024-03-19"
output:
  word_document: default
  pdf_document: default
---

# Q1
```{r}
x1 <- c(574, 976, 789, 805, 361, 529)
x2 <- c(791, 1146, 394, 767, 1385, 1021, 2073, 803, 1263, 1016, 1101, 945, 139)
x3 <- c(865, 775, 729, 1721, 1113, 820, 1613, 1404, 1201, 205, 380, 580, 1803)
x4 <- c(998, 1049, 736, 782, 730, 742, 1219, 705, 1260, 611, 1350, 1657, 1143)
x5 <- c(1154, 541, 406, 1529, 1132, 767, 1224, 314, 1728)
grps = rep(1:5, c(6, 13, 13, 13, 9))
x <- c(x1, x2, x3, x4, x5)

getmeans <- function(x,grps)
{
  #estimate the group-specific means
  junk <- table(grps)
  meanvec <- rep(NA,length(junk))
  for (i in 1:length(junk)) meanvec[i] <- mean(x[grps==names(junk)[i]])
  meanvec
}

getF <- function(x,grps)
{
  # estimate the F test statistic for one-way layout
  junk <- table(grps)
  SST <- sum(junk * (getmeans(x,grps))^2) -length(x)*(mean(x))^2
  SST / (length(junk)-1) / ( (sum((x-mean(x))^2) - SST)/(length(x)-length(junk)))
}

perm.approx.F <- function(x,grps,R)
{
  #x: observed response vector
  #grps: group number
  #R: number of permutations
  results <- rep(NA,R)
  #for (i in 1:R) results[i] <- summary(aov(x[sample(1:(length(x)),length(x))] ~ factor(grps)))[1,4]
  for (i in 1:R) results[i] <- getF(x[sample(1:(length(x)),length(x))],grps)
  results
}

#(trtmeans <- getmeans(x,grps))

# Observed F-value
#Fobs <- summary(aov(x~factor(grps)))[[1]][1,4]
Fobs = getF(x,grps)
perm.F = perm.approx.F(x, grps, R=1000)
(perm.pval = mean(perm.F >= Fobs))

#ANOVA based on the assumption of normal distribution with equal variance
summary(aov(x ~ factor(grps)))

```
From the results, we can see that the p-value of permutation F test is 0.549, and the p-value of ANOVA F test is 0.557.


# Q2
```{r}
x1 <- c(0, 2, 1, 3, 1, 1)
x2 <- c(2, 0, 3, 1, 3, 4)
x3 <- c(1, 3, 4, 2, 2, 1)
x4 <- c(3, 4, 2, 5, 3, 4)
x <- c(x1, x2, x3, x4)
grps <- rep(1:4, each = 6)

rank.x = rank(x)

# use f test to calculate KW statistics
N = length(x)
SST = summary(aov(rank.x ~ factor(grps)))[[1]][1,2]
KW <- SST*12/(N*(N+1))
KW


# use equation to calculate KW statistics
#sum <- 0
#means = getmeans(rank.x, grps)
#for (i in 1:4) {
#  sum <- sum + (means[i]-(N+1)/2)^2 * 6
#}
#KW <- 12 * sum /(N * (N+1))

# Kruskal-Wallis test with chi-square approximation
#summary(aov(rank.x ~ factor(grps)))
#kruskal.test(x,grps)



# K-W test with permutation-based approx p-value
Fobs <- getF(rank.x, grps)
set.seed(123)
permFs <- perm.approx.F(rank.x, grps, R=1000)
mean(permFs >= Fobs)
```
The KW statistics is 7.295. Use permutation method to calculate the approximate p-value and the p-value is 0.054 > 0.05. We can not reject $H_0$

# Q3
```{r}
library(clinfun)
x1 <- c(574, 976, 789, 805, 361, 529)
x2 <- c(791, 1146, 394, 767, 1385, 1021, 2073, 803, 1263, 1016, 1101, 945, 139)
x3 <- c(865, 775, 729, 1721, 1113, 820, 1613, 1404, 1201, 205, 380, 580, 1803)
x4 <- c(998, 1049, 736, 782, 730, 742, 1219, 705, 1260, 611, 1350, 1657, 1143)
x5 <- c(1154, 541, 406, 1529, 1132, 767, 1224, 314, 1728)
x <- c(x1, x2, x3, x4, x5)
grps <- rep(1:5, c(6, 13, 13, 13, 9))
jonckheere.test(x, grps, alternative="increasing") #based on MW test statistic
```
The p-value of JT test is 0.181.

# Q4
```{r}
x1 <- c(49.17, 48.19, 48.93)
x2 <- c(4.48, 4.43, 4.79)
x3 <- c(6.36, 7.03, 5.36)
x4 <- c(5.34, 4.86, 6.64)
x5 <- c(5.17, 6.47, 6.58)
x6 <- c(3.99, 5.89, 6.61)
x7 <- c(6.02, 4.96, 6.30)
x8 <- c(3.98, 6.30, 4.96)
x9 <- c(5.94, 6.03, 5.56)
x10 <- c(6.29, 5.11, 5.93)

x <- c(x1, x2, x3, x4, x5, x6, x7, x8, x9, x10)
grps <- rep(1:10, each = 3)
xmeans <- getmeans(x, grps)

perm.approx.diff <- function(x,n1,n2,R)
{
### approximate permutation of pairwise group means
### n1: sample size of group 1
### n2: sample size of group 2
### x: combined data
### R: number of random permutations
### assumption: the observations in x are exchangeable under H0
    results <- rep(NA,R)
    for (i in 1:R)
    {
        temp <- x[sample(1:(length(x)),n1+n2)]
        results[i] <- mean(temp[1:n1]) - mean(temp[(n1+1):(n1+n2)])
    }
    results
}

Fisher.LSD = function(x, grps, k, alpha=0.05, R=1000)
{
    #Fisher's Protected LSD
    ##
    ### overall test (permutation F-test)
    Fobs = getF(x, grps)
    perm.F = perm.approx.F(x, grps, R=R)
    perm.pval = mean(perm.F >= Fobs)
   nn <- table(factor(grps))

    if(perm.pval<=alpha)
    {
        cat("the overall test is significant, proceed with pairwise comparison", "\n")

        unadj.pvals <- matrix(NA,k,k)
        for (i in 2:k) {
            for (j in 1:(i-1)){
                diff.perms = perm.approx.diff(x[grps %in% c(i, j)], nn[i], nn[j], R=R)
                diff.obs = mean(x[grps==i]) - mean(x[grps==j])
                unadj.pvals[i,j] <- mean(abs(diff.perms) >= abs(diff.obs))
          }}
        sig = (unadj.pvals <= alpha)
        out = list(pval.overall= perm.pval, pval.pairwise= unadj.pvals, sig=sig)
    }
    else if(perm.pval>alpha)
    {
        cat("the overall test is not significant, do not proceed with pairwise comparison", "\n")
        out = list(pval.overall= perm.pval)
        }
    return(out)
}

k <- length(xmeans)
Fisher.LSD(x, grps, k, alpha = 0.05, R = 1000)

```
## (a)
```{r}
library(agricolae)
x1 <- c(49.17, 48.19, 48.93)
x2 <- c(4.48, 4.43, 4.79)
x3 <- c(6.36, 7.03, 5.36)
x4 <- c(5.34, 4.86, 6.64)
x5 <- c(5.17, 6.47, 6.58)
x6 <- c(3.99, 5.89, 6.61)
x7 <- c(6.02, 4.96, 6.30)
x8 <- c(3.98, 6.30, 4.96)
x9 <- c(5.94, 6.03, 5.56)
x10 <- c(6.29, 5.11, 5.93)

x <- c(x1, x2, x3, x4, x5, x6, x7, x8, x9, x10)
grps <- rep(1:10, each = 3)
xmeans <- getmeans(x, grps)
fit <- aov(x~factor(grps))

LSD.test(fit, "factor(grps)", console = T)
```
From the results, we can see that LSD can identify differences between treatment 1 and the rest. But it also identifies the difference between treatment 2 and treatment 3-10, although actually they have the same $\mu$.

## (b)

```{r}
Bonf.adj = function(x, grps, k, alpha, R=1000, test=c("perm.meandiff", "ttest"))
{
    #Bonferroni correction
    #alpha: overall significance level
    #R: number of permutation
    #test="perm.meandiff": pvalues are calculated through permutation test based on mean difference from two groups
    #test="ttest": pvalues are calculated through pairwise two-sample t-test (requires normal distirbution assumption)

    #output:
    #adj.alpha: adjsuted significance level for each pairwise comparison
    # bonf.pvals: p-val from pairwise comparison
    # sig: TRUE means the two pairs are significantly different

    nn <- table(factor(grps))
    ## obtain the permutation p-value for comparing each pair (i, j)
    bonf.pvals = matrix(NA, k, k)
    if(test=="perm.meandiff")
    {
    for (i in 2:k) {
        for (j in 1:(i-1)) {
            diff.perms = perm.approx.diff(x[grps %in% c(i, j)], nn[i], nn[j], R=R)
            diff.obs = mean(x[grps==i]) - mean(x[grps==j])
            bonf.pvals[i,j] <- mean(abs(diff.perms) >= abs(diff.obs))
          }}
    }

    if(test=="ttest")
    {
        for (i in 2:k) {
        for (j in 1:(i-1)) {
            bonf.pvals[i,j] = t.test(x[grps==i], x[grps==j])$p.value
          }}
    }
    # For Bonferroni, compare each of the p-values against the adjusted level: alpha/(k(k-1)/2)
    (adj.alpha = alpha / (k*(k-1)/2))
    sig = (bonf.pvals <= adj.alpha)
    out = list(adj.alpha=adj.alpha, bonf.pvals=bonf.pvals, sig=sig)
    return(out)
}

set.seed(123)  #you can change the seed
Bonf.adj(x, grps, k, alpha=0.05, R=1000, test="perm.meandiff")
Bonf.adj(x, grps, k, alpha=0.05, R=1000, test="ttest")

```

```{r}

getmaxTij <- function(x, grps, MSE)
{
    # estimate the maximum of Tij (pairwise mean diff) of a given data x
    trtmeans <- getmeans(x,grps)
   nn <- table(factor(grps))
    k <- length(trtmeans)
    Tijs <- matrix(NA,k,k)
    for (i in 2:k) {
    for (j in 1:(i-1)){
         Tijs[i,j] <- abs(trtmeans[i] - trtmeans[j])/sqrt(MSE/2 * (1/nn[i] + 1/nn[j]))
     }}
    max(Tijs,na.rm=T)
}

perm.approx.maxTij <- function(x,grps,MSE,R)
{
    ### obtain the null permutation distribution of maxTij
    results <- rep(NA,R)
    for (i in 1:R) results[i] <- getmaxTij(x[sample(1:(length(x)),length(x))],grps,MSE)
    results
}

Tukey.HSD = function(x, grps, k, alpha=0.05, R=1000)
{
    #Tukey's HSD
    #summary(aov(x ~ factor(grps)))
    nn <- table(factor(grps))
    trtmeans <- getmeans(x,grps)

    (MSE <- summary(aov(x ~ factor(grps)))[[1]][2,3])
    ### observed Tij
    Tijs <- matrix(NA,k,k)
    for (i in 2:k){
        for (j in 1:(i-1)){
             Tijs[i,j] <- abs(trtmeans[i] - trtmeans[j])/sqrt(MSE/2 * (1/nn[i] + 1/nn[j]))
             }}

    ### observed maxTij
    #getmaxTij(x,grps,MSE)

    ### permutation maxTij
    perm.maxTij <- perm.approx.maxTij(x,grps,MSE,R)

    pvalsTij <- matrix(NA,k,k)
    for (i in 2:k){
    for (j in 1:(i-1)){
        pvalsTij[i,j] <- mean(perm.maxTij >= Tijs[i,j])
        }}

    ### compare the pairwise pvalue with alpha
    sig = (pvalsTij <= alpha)

    out = list(sig=sig, pvalsTij= pvalsTij)
    return(out)
}
Tukey.HSD(x, grps, k, alpha=0.05, R=1000)
```
Bonferroni method and Tukey's HSD can not recognize the differences among the means of treatments 2 through 10.

# Q5
```{r}
set.seed(123)
tr1 <- c(100, 250, 50, 80)
tr2 <- c(112, 240, 58, 82)
d <- tr2 - tr1
R <- 10000
srplus <- rep(NA, R)
for (i in 1:R){
  sign <- sample(c(0, 1), 4, replace = TRUE)
  srplus[i] <- sign %*% abs(d)
}

a <- table(srplus)
prop.table(a)
```



# Q6
## (a)
```{r}
x1 <- c(1.5, 2.1, 1.9, 2.8, 1.4, 1.8)
x2 <- c(1.8, 2, 2, 2.7, 1.6, 2.3)
x3 <- c(1.9, 2.5, 2.5, 2.6, 2.1, 2.4)
x <- c(x1, x2, x3)
blocks = rep(1:6,3)
grps = rep(1:3, each = 6)

#F-test
summary(aov(x~factor(grps)+factor(blocks)))


#  Permutation Test for RCBD
### Step 1: calculate observed SSTM:sum of squared group-means

### calculate observed SSTM
getSSTM <- function(x,grps)
{
    ## estimate SSTM=sum of squared group-means 
    sum( (getmeans(x,grps))^2 )
    }

### obtain the permutation distribution of SSTM under H0
###             i.e. obtain SSTM*'s with R permutations
perm.approx.RCBD <- function(x,grps,blocks,R)
{
    # obtain the aprpoximate permutation distribution of SSTM under the null hypothesis
    mat <- cbind(x,grps,blocks)
    mat <- mat[order(mat[,3]),] 
    k <- length(table(grps))
    b <- length(table(blocks))
    results <- rep(NA,R)
    for (i in 1:R)
    {
        junk <- rep(NA,b*k)
        for (j in 1:b) junk[k*(j-1)+(1:k)] <- sample(1:k,k)
        results[i] <- getSSTM(mat[,1],junk)
    }
    results
}
SSTMobs <- getSSTM(x,grps)
SSTMobs

### Step 2: obtain the permutation distribution of SSTM under H0
###             i.e. obtain SSTM*'s with R permutations

permSSTM <- perm.approx.RCBD(x,grps,blocks,R=1000)

### Step3: calculate the p-value
(pval = mean(permSSTM >= SSTMobs))
```

## (b)
```{r}
summary(aov(x~blocks + grps))
```
ANOVA can not recogni the difference between treatments.

